---
title: "PS2 Differential Abundance w/ DESeq2"
output: html_notebook
---

This notebook was used for taxonomic association analyses. Before executing it, be sure to change the file paths in "File Import & Phyloseq Conversion" to reflect your own file paths. Some portions of the code have been commented out; optionally, you can un-comment them. 


```{r}
library(phyloseq)
library(genefilter)
library(ggplot2)
library(ggpubr)
library(reshape2)
library(RColorBrewer)
library(viridis)
library(wesanderson)
library(grid)
library(gridExtra)
library(plotly)
library(scales)
library(dplyr)
set.seed(1)
```

# File Import & Phyloseq Conversion
```{r}
# Import biom table, tree, and mapping
biom_file <- "/path/to/biom"
tree_file <- "/path/to/tree"
map_file <- "/path/to/mapping"


biom_otu_tax <- import_biom(biom_file, tree_file, parseFunction=parse_taxonomy_default)
biom_otu_tax
```

```{r}
# Rename taxonomic levels
colnames(tax_table(biom_otu_tax)) <- c(Rank1="Domain",Rank2="Phylum",Rank3="Class",Rank4="Order",Rank5="Family",Rank6="Genus",Rank7="Species")
```

```{r}
# Add sample data to the dataset w/ merge
bmsd <- import_qiime_sample_data(map_file)
dim(bmsd)
ps2 <- merge_phyloseq(biom_otu_tax, bmsd, tree_file)
ps2
```
```{r}
#  Remove the SILVA prefix "D_#__" from the taxonomy
tax_table(ps2) <- gsub("D_[0-9]__","", tax_table(ps2)) # use gsub to remove
#tax_table(ps2)[1:5, 1:7] # preview the edited taxons
```
```{r}
# root the tree
tree.unrooted <- phy_tree(ps2)
pick_new_outgroup <- function(tree.unrooted){
    require("magrittr")
    require("data.table")
    require("ape") # ape::Ntip
    # tablify parts of tree that we need.
    treeDT <- 
      cbind(
        data.table(tree.unrooted$edge),
        data.table(length = tree.unrooted$edge.length)
      )[1:Ntip(tree.unrooted)] %>% 
      cbind(data.table(id = tree.unrooted$tip.label))
    # Take the longest terminal branch as outgroup
    new.outgroup <- treeDT[which.max(length)]$id
    return(new.outgroup)
  }
new.outgroup = pick_new_outgroup(tree.unrooted)
rootedTree = ape::root(tree.unrooted, outgroup=new.outgroup, resolve.root=TRUE)
phy_tree(ps2) <- rootedTree
phy_tree(ps2)
```
```{r}
# PRE-PROCESSING

# pre-filter the table to remove controls
ps2.pf <- ps2
#ps2.pf <- subset_samples(ps2, !(Patient %in% c("15","16","18")))
#ps2.pf <- subset_samples(ps2.pf, !(SampleID %in% c("35A","35B","38A")))
#ps2.pf <- subset_samples(ps2.pf, SampleType %in% c("Pre-Debridement","Post-Debridement","Skin"))
#ps2.pf = prune_taxa(taxa_sums(ps2.pf) >= 1, ps2.pf)
```


#```{r}
# Genus agglomeration (should this be tip_glom? idk anymore...)
ps2.pf.gg <- tax_glom(ps2.pf, "Genus", NArm = FALSE)
ps2.pf.ggra <- transform_sample_counts(ps2.pf.gg, function(x) (x / sum(x))*100 )

# Pre-melt (Replace other indiv. melts w/ these)
ps2.pf.gg.melt <- psmelt(ps2.pf.gg)
ps2.pf.ggra.melt <- psmelt(ps2.pf.ggra)
#```

# DESeq Analyses
```{r}
# DESeq2 differential abundance analysis
# Overall comparisons

library(DESeq2)

# These lines will produce the original DESeq2 object, and run the first comparison 
# convert to DESeq2
ps2.dds = phyloseq_to_deseq2(ps2.pf, ~ 1)
# refactor levels for desired comparisons
ps2.dds$SampleType <- factor(ps2.dds$SampleType, levels = c("Pre-Debridement","Post-Debridement","Skin"))
# specify experiment design
design(ps2.dds) <- ~ SampleType
# run DESeq
dds.out = DESeq(ps2.dds, test="Wald", fitType="parametric")

# Pre vs. Skin Results
res.ps = results(dds.out, contrast = c("SampleType","Pre-Debridement","Skin"), cooksCutoff = FALSE)
write.csv(res.ps, "pre.vs.skin.csv")

# Pre vs. Post Overall Results
res.pp = results(dds.out, contrast = c("SampleType","Pre-Debridement","Post-Debridement"), cooksCutoff = FALSE)
write.csv(res.pp, "pre.vs.post.overall.csv")

# Wound (Pre + Post) vs. Skin DESeq & Results
design(ps2.dds) <- ~ Description
dds.out = DESeq(ps2.dds, test="Wald", fitType="parametric")
res.ws = results(dds.out, contrast = c("Description","Wound","Skin"), cooksCutoff = FALSE)
write.csv(res.ws, "wound.vs.skin.overall.csv")

# Pre vs. Post HEALED ONLY DESeq & Results
# Create nested group & run DESeq
ps2.dds$group <- factor(paste0(ps2.dds$Outcome, ps2.dds$prepost))
design(ps2.dds) <- ~ group
dds.hu = DESeq(ps2.dds, test="Wald", fitType="parametric")
res.hpp = results(dds.hu, contrast = c("group","HealedPre","HealedPost"), cooksCutoff = FALSE)
write.csv(res.hpp, "pre.vs.post.healed.csv")

# Healed vs. Unhealed (Post- Only) Results
res.hupost = results(dds.hu, contrast = c("group","HealedPost","UnhealedPost"), cooksCutoff = FALSE)
write.csv(res.hupost, "healed.vs.unhealed.post.csv")

# Pre vs. Post UNHEALED ONLY DESeq & Results
# Create nested group & run DESeq
ps2.dds$group <- factor(paste0(ps2.dds$Outcome, ps2.dds$prepost))
# Refactor to get desired comparisons
ps2.dds$group <- factor(ps2.dds$group, levels = c("UnhealedPre","UnhealedPost","HealedPre","HealedPost","SkinControl"))
design(ps2.dds) <- ~ group
dds.hu = DESeq(ps2.dds, test="Wald", fitType="parametric")
res.upp = results(dds.hu, contrast = c("group","UnhealedPre","UnhealedPost"), cooksCutoff = FALSE)
write.csv(res.upp, "pre.vs.post.unhealed.csv")

# Healed vs. Unhealed (Overall) DESeq & Results
design(ps2.dds) <- ~ Outcome
dds.out = DESeq(ps2.dds, test="Wald", fitType="parametric")
res.hu = results(dds.out, contrast = c("Outcome","Healed","Unhealed"), cooksCutoff = FALSE)
write.csv(res.hu, "healed.vs.unhealed.overall.csv")

# Healed vs. Unhealed (Pre- Only) Results
res.hupre = results(dds.hu, contrast = c("group","HealedPre","UnhealedPre"), cooksCutoff = FALSE)
write.csv(res.hupre, "healed.vs.unhealed.pre.csv")
```
# Significant OTU Visualizations
## Pre vs. Skin
```{r}
alpha = 0.05
sigtab = res.ps[which(res.ps$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(ps2.pf)[rownames(sigtab), ], "matrix"))
head(sigtab)
dim(sigtab)
x = tapply(sigtab$log2FoldChange, sigtab$Phylum, function(x) max(x))
x = sort(x, TRUE)
sigtab$Phylum = factor(as.character(sigtab$Phylum), levels=names(x))

# Genus order
x = tapply(sigtab$log2FoldChange, sigtab$Genus, function(x) max(x))
x = sort(x, TRUE)
sigtab$Genus = factor(as.character(sigtab$Genus), levels=names(x))

# plot
p1 <- ggplot(sigtab, aes(y=Genus, x=log2FoldChange, color=Phylum)) + 
  geom_point(size=4) + theme_bw()
  #theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
p1
png("pre.vs.skin.png", width = 6, height = 12, units = "in", res = 600)
p1
dev.off()
```

## Wounds (Pre + Post) vs. Skin
```{r}
alpha = 0.05
sigtab = res.ws[which(res.ws$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(ps2.pf)[rownames(sigtab), ], "matrix"))
head(sigtab)
dim(sigtab)
x = tapply(sigtab$log2FoldChange, sigtab$Phylum, function(x) max(x))
x = sort(x, TRUE)
sigtab$Phylum = factor(as.character(sigtab$Phylum), levels=names(x))

# Genus order
x = tapply(sigtab$log2FoldChange, sigtab$Genus, function(x) max(x))
x = sort(x, TRUE)
sigtab$Genus = factor(as.character(sigtab$Genus), levels=names(x))

# plot
p1 <- ggplot(sigtab, aes(y=Genus, x=log2FoldChange, color=Phylum)) + 
  geom_point(size=4) + theme_bw()
  #theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
p1
png("wound.vs.skin.overall.png", width = 6, height = 12, units = "in", res = 600)
p1
dev.off()
```

## Pre vs. Post (Overall)
```{r}
# Pre vs. Post Overall
alpha = 0.05
sigtab = res.pp[which(res.pp$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(ps2.pf)[rownames(sigtab), ], "matrix"))
head(sigtab)
dim(sigtab)
x = tapply(sigtab$log2FoldChange, sigtab$Phylum, function(x) max(x))
x = sort(x, TRUE)
sigtab$Phylum = factor(as.character(sigtab$Phylum), levels=names(x))

# Genus order
x = tapply(sigtab$log2FoldChange, sigtab$Genus, function(x) max(x))
x = sort(x, TRUE)
sigtab$Genus = factor(as.character(sigtab$Genus), levels=names(x))

# plot
# 
p1 <- ggplot(sigtab, aes(y=Genus, x=log2FoldChange, color=Phylum)) + 
  geom_point(size=4) + theme_bw()
  #theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
p1
png("pre.vs.post.overall.png", width = 6, height = 12, units = "in", res = 600)
p1
dev.off()
```

## Pre vs. Post (Healed Only)
```{r}
alpha = 0.05
sigtab = res.hpp[which(res.hpp$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(ps2.pf)[rownames(sigtab), ], "matrix"))
head(sigtab)
dim(sigtab)
x = tapply(sigtab$log2FoldChange, sigtab$Phylum, function(x) max(x))
x = sort(x, TRUE)
sigtab$Phylum = factor(as.character(sigtab$Phylum), levels=names(x))

# Genus order
x = tapply(sigtab$log2FoldChange, sigtab$Genus, function(x) max(x))
x = sort(x, TRUE)
sigtab$Genus = factor(as.character(sigtab$Genus), levels=names(x))

# plot
# 
p1 <- ggplot(sigtab, aes(y=Genus, x=log2FoldChange, color=Phylum)) + 
  geom_point(size=4) + theme_bw()
  #theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
p1
png("pre.vs.post.healed.png", width = 6, height = 12, units = "in", res = 600)
p1
dev.off()
```

## Pre vs. Post (Unhealed Only)
```{r}
alpha = 0.05
sigtab = res.upp[which(res.upp$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(ps2.pf)[rownames(sigtab), ], "matrix"))
head(sigtab)
dim(sigtab)
x = tapply(sigtab$log2FoldChange, sigtab$Phylum, function(x) max(x))
x = sort(x, TRUE)
sigtab$Phylum = factor(as.character(sigtab$Phylum), levels=names(x))

# Genus order
x = tapply(sigtab$log2FoldChange, sigtab$Genus, function(x) max(x))
x = sort(x, TRUE)
sigtab$Genus = factor(as.character(sigtab$Genus), levels=names(x))

# plot
# 
p1 <- ggplot(sigtab, aes(y=Genus, x=log2FoldChange, color=Phylum)) + 
  geom_point(size=4) + theme_bw()
  #theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
p1
png("pre.vs.post.unhealed.png", width = 6, height = 12, units = "in", res = 600)
p1
dev.off()
```

## Healed vs. Unhealed (Overall)
```{r}
# Healed vs. Unhealed Overall
alpha = 0.05
sigtab = res.hu[which(res.hu$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(ps2.pf)[rownames(sigtab), ], "matrix"))
head(sigtab)
dim(sigtab)
x = tapply(sigtab$log2FoldChange, sigtab$Phylum, function(x) max(x))
x = sort(x, TRUE)
sigtab$Phylum = factor(as.character(sigtab$Phylum), levels=names(x))

# Genus order
x = tapply(sigtab$log2FoldChange, sigtab$Genus, function(x) max(x))
x = sort(x, TRUE)
sigtab$Genus = factor(as.character(sigtab$Genus), levels=names(x))

# plot
# 
p1 <- ggplot(sigtab, aes(y=Genus, x=log2FoldChange, color=Phylum)) + 
  geom_point(size=4) + theme_bw()
  #theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
p1
png("healed.vs.unhealed.overall.png", width = 6, height = 12, units = "in", res = 600)
p1
dev.off()
```

## Healed vs. Unhealed (Pre- Only)
```{r}
# Healed vs. Unhealed Pre- only
alpha = 0.05
sigtab = res.hupre[which(res.hupre$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(ps2.pf)[rownames(sigtab), ], "matrix"))
head(sigtab)
dim(sigtab)
x = tapply(sigtab$log2FoldChange, sigtab$Phylum, function(x) max(x))
x = sort(x, TRUE)
sigtab$Phylum = factor(as.character(sigtab$Phylum), levels=names(x))

# Genus order
x = tapply(sigtab$log2FoldChange, sigtab$Genus, function(x) max(x))
x = sort(x, TRUE)
sigtab$Genus = factor(as.character(sigtab$Genus), levels=names(x))

# plot
# 
p1 <- ggplot(sigtab, aes(y=Genus, x=log2FoldChange, color=Phylum)) + 
  geom_point(size=4) + theme_bw()
  #theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
p1
png("healed.vs.unhealed.pre.png", width = 6, height = 12, units = "in", res = 600)
p1
dev.off()
```

## Healed vs Unhealed (Post- Only)
```{r}
# Healed vs. Unhealed Post- only
alpha = 0.05
sigtab = res.hupost[which(res.hupost$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(ps2.pf)[rownames(sigtab), ], "matrix"))
head(sigtab)
dim(sigtab)
x = tapply(sigtab$log2FoldChange, sigtab$Phylum, function(x) max(x))
x = sort(x, TRUE)
sigtab$Phylum = factor(as.character(sigtab$Phylum), levels=names(x))

# Genus order
x = tapply(sigtab$log2FoldChange, sigtab$Genus, function(x) max(x))
x = sort(x, TRUE)
sigtab$Genus = factor(as.character(sigtab$Genus), levels=names(x))

# plot
# 
p1 <- ggplot(sigtab, aes(y=Genus, x=log2FoldChange, color=Phylum)) + 
  geom_point(size=4) + theme_bw()
  #theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
p1
png("healed.vs.unhealed.post.png", width = 6, height = 12, units = "in", res = 600)
p1
dev.off()
```